preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-03 Manuscript/manuscript.Rmd
Minor edits
completed todo items

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state data (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://dozmorovlab.github.io/preciseTAD/ and Bioconductor (submitted).

@@

To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold $t$ specifies the probability at which a base with $p_{n_{i}} \ge t$ is designated as boundary (the default $t=1$). Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between boundary-annotated bases. The resulting clusters of spatially colocalized bases predicted as being boundaries were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around medoids clustering (CLARA) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods).

@@

In summary, we demonstrate that domain boundary prediction is a multi-faceted problem requiring consideration of multiple statistical and biological properties of genomic data. Simply considering the properties of Hi-C contact matrices ignores the fundamental roles of known molecular drivers of the 3D chromatin structures. Instead, we propose *preciseTAD*, a supervised machine learning framework that leverages both Hi-C contact matrix information and genomic annotations. Our method introduces three concepts - *shifted binning*, *distance*-type predictors, and *random undersampling* - which we use to build random forest classification models for predicting boundary regions. Our method can bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. We introduce *preciseTAD*, an open source R package for leveraging random forests to predict domain boundaries at base-думуд resolution. We hope that *preciseTAD* will serve as an efficient and easy-to-use tool to further explore the genome's 3D organization.

@@

AUC - Area Under the Curve; BA - Balance Accuracy; bp - base pair; BroadHMM - BroadHMM chromatin state data; ChIP-seq - Chromatin Immunoprecipitation followed by high-throughput DNA sequencing; CI - Class Imbalance; CLARA - Clustering Large Applications; DBSCAN - Density-Based Spatial Clustering of Applications with Noise; FN - False Negatives; FP - False Positives; Hi-C - High-throughput chromosome conformation capture technology; HM - Histone Modifications; kb - kilobase; ML - Machine Learning; OC - Overlap Counts; OP - Overlap Percents; PAM - Partitioning Around Medoids; PTBP - preciseTAD Boundary Point; PTBR - preciseTAD Boundary Region; RF - Random Forest; RFE - Recursive Feature Elimination; ROC - Receiver Operating Characteristic; ROS - Random Over-Sampling; RUS	Random Under-Sampling; SMOTE - Synthetic Minority Over-sampling Technique; TAD - Topologically Associating Domains; TFBS - Transcription Factor Binding Sites; TN - True Negatives; TP - True Positives

>>>>

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state data (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://dozmorovlab.github.io/preciseTAD/ and Bioconductor (submitted).

@@

To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold $t$ specifies the probability at which a base with $p_{n_{i}} \ge t$ is designated as boundary (the default $t=1$). Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between boundary-annotated bases. The resulting clusters of spatially colocalized bases predicted as being boundaries were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements partitioning around medoids (PAM) on the sub distance matrix derived from each cluster. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base-level resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with a genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict each base's probability of being a boundary. Our method, referred to as *preciseTAD*, uses density-based spatial clustering (DBSCAN) and partitioning around medoids (PAM) to predict the location of boundary regions and summit points (highest boundary probability) at base-level resolution (see Methods).

@@

In summary, we demonstrate that domain boundary prediction is a multi-faceted problem requiring consideration of multiple statistical and biological properties of genomic data. Simply considering the properties of Hi-C contact matrices ignores the fundamental roles of known molecular drivers of the 3D chromatin structures. Instead, we propose *preciseTAD*, a supervised machine learning framework that leverages both Hi-C contact matrix information and genomic annotations. Our method introduces three concepts - *shifted binning*, *distance*-type predictors, and *random undersampling* - which we use to build random forest classification models for predicting boundary regions. Our method can bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. We introduce *preciseTAD*, an open source R package for leveraging random forests to predict domain boundaries at base-level resolution. We hope that *preciseTAD* will serve as an efficient and easy-to-use tool to further explore the genome's 3D organization.

@@

AUC - Area Under the Curve; BA - Balance Accuracy; bp - base pair; BroadHMM - BroadHMM chromatin state data; ChIP-seq - Chromatin Immunoprecipitation followed by high-throughput DNA sequencing; CI - Class Imbalance; DBSCAN - Density-Based Spatial Clustering of Applications with Noise; FN - False Negatives; FP - False Positives; Hi-C - High-throughput chromosome conformation capture technology; HM - Histone Modifications; kb - kilobase; ML - Machine Learning; OC - Overlap Counts; OP - Overlap Percents; PAM - Partitioning Around Medoids; PTBP - preciseTAD Boundary Point; PTBR - preciseTAD Boundary Region; RF - Random Forest; RFE - Recursive Feature Elimination; ROC - Receiver Operating Characteristic; ROS - Random Over-Sampling; RUS	Random Under-Sampling; SMOTE - Synthetic Minority Over-sampling Technique; TAD - Topologically Associating Domains; TFBS - Transcription Factor Binding Sites; TN - True Negatives; TP - True Positives
